Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-04-27 ms.tex
cutting fat - and lots of it3

We first describe analytic estimates of physical quantities, such as
the stagnation radius $\rs$ and the mass accretion rate, the detailed
derivation of which are given in Appendix \ref{app:rs}.  Our results
are expressed in terms of the SMBH mass $M_{\bullet}$, total heating
rate $\tilde{v}_{w}$, and wind mass loss parameter $\eta$.

@@

The accretion rate towards the SMBH is given by the total mass loss rate interior to the stagnation radius (eq.~[\ref{eq:q}]), 
\begin{eqnarray}
  \dot{M} &=& 4\pi \int_{0}^{r_{s}}q r^{2}dr = \frac{\eta M_{\star}|_{r_{\rm s}}}{\th} \nonumber \\
&\approx&
  \begin{cases}
    4.5 \times 10^{-5} M_{\bullet,8}^{1.76}
    v_{500}^{-3.8}  \eta_{0.02} \Msun \, \pyear& \text{core} \\
    3.2 \times 10^{-5} M_{\bullet,8}^{1.48} 
    v_{500}^{-2.4}  \eta_{0.02} \Msun \, \pyear  & \text{cusp}, 
  \end{cases}
  \label{eq:mdot_analytic}
\end{eqnarray}
where we have assumed $v_{w} \gg \sigma_0$ by adopting equation
(\ref{eq:stag_simple}) for $r_s$.  The resulting Eddington ratio is
given by
\begin{eqnarray}
\frac{\dot{M}}{\dot{M}_{\rm edd}} &\approx&
  \begin{cases}
    2.0 \times 10^{-5} M_{\bullet,8}^{0.76}
    v_{500}^{-3.8}  \eta_{0.02}   & \text{core}, \\
    1.4 \times 10^{-5} \Mbheight^{0.48} 
    v_{500}^{-2.4}  \eta_{0.02}   & \text{cusp}, 
  \end{cases}
  \label{eq:eddr_analytic}
\end{eqnarray}
where $\dot{M}_{\rm edd} = 2.6M_{\bullet,8}M_{\odot}$ yr$^{-1}$ is the
Eddington accretion rate, assuming a radiative efficiency of ten per
cent.  Note the sensitive dependence of the accretion rate on the wind
heating rate.  We emphasize that equation~\eqref{eq:eddr_analytic} represents the radial mass inflow rate on relatively large scales and does not include outflows from the accretion disk on small scales (e.g.~\citealt{Blandford&Begelman99}; \citealt{Li+13}), which could significantly reduce the fraction of $\dot{M}$ actually reaching the SMBH.

@@

Our analytic estimates accurately reproduce the numerical results in
the high heating limit $\zeta \gg 1$ ($v_w \gg \sigma_0$; $r_{\rm s}
\lesssim r_{\rm inf}$).  However, for low heating the stagnation
radius diverges above the analytic estimate, approaching the stellar
break radius $r_b \gg r_{\rm inf}$.  This divergence occurs
approximately when $\zeta < \zeta_c \propto (r_b/r_{\rm
  soi})^{0.5(\Gamma-1)}$ (eq.~[\ref{eq:zetacrit}]), which coincides
with the Bernoulli parameter of the gas decreasing below unity near
the stagnation radius (Appendix \ref{app:be}).  Thus, for small values
of the heating rate (small $\zeta$) the location of the stagnation
radius will vary strongly with the break radius. This explains the
behavior of the three vertically aligned green squares. These are three
cusp ($\Gamma=0.8$) galaxies with $\Mbh=10^8 \Msun$ and $v_w=300$ km/s
but with different break radii ($\rb=$ 25, 50, and 100 pc from top to
bottom). This divergence of the stagnation radius to large radii
occurs at a higher value of $\zeta$ in core galaxies ($\Gamma = 0.1$),
explaining the behavior of the two core galaxies shown in
Fig.~\ref{fig:stag} as vertically aligned orange triangles.  These
solutions correspond to the same black hole mass ($\Mbh = 10^8 \Msun$)
and heating rate ($v_w = 600$ km s$^{-1}$) but assume different break
radii of $r_{b} =$ 25, 50 pc (from top to bottom).

@@

Also note that for $v_{w}=600$ and $1200$ km s$^{-1}$ we have that
$\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}| \sim t_{\rm cool}/t_{\rm ff}$
near the stagnation radius.  However, the two ratios diverge from each
other at low heating ($v_{w} \lesssim 300$ km s$^{-1}$) because
equations (\ref{eq:mdot_gas}), (\ref{eq:rhors}) underestimate the true
gas density, which is greater in a subsonic flow (weak heating) than
in a supersonic outflow (strong heating).

In addition to our solutions which neglect thermal conduction and
radiative cooling, we explore a smaller set of simulations including
these effects explicitly, as marked by dagger symbols in Table
\ref{table:models}.  For solutions which are far from being thermally
unstable when cooling is neglected (e.g., $M_{\bullet} =
10^{8}M_{\odot}$, $v_{w} = 600$ km s$^{-1}$, $r_{b} = 100$ pc, $\eta =
0.2$, $\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}| = 76$), the inclusion of
radiative cooling has little effect on the key properties of the flow,
e.g., stagnation radius, mass accretion rate, $\dot{q}_{\rm
  heat}/|\dot{q}_{\rm rad}|$.  However, for solutions that are only
marginally thermally stable, including radiative losses acts to
decrease the value of $\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}|$ by a
greater fractional amount.  We calculated solutions with radiative
cooling for all cusp galaxies with $v_w=300$ km s$^{-1}$ and for all
core galaxies with $v_w=600 $ km s$^{-1}$, each for three different
values of $\eta = 0.02, 0.2, 0.6$, spanning the physical range of
expected mass loss for continuous star formation (see
Fig.~\ref{fig:vwSources}).  Solutions found to be unstable for all
values of $\eta$ are marked in boldface as ``TI" in Table
\ref{table:models}, with the unstable values of $\eta$ provided in the
final column.

Thermal conduction, by contrast, results in only minor changes to our
solutions for our fiducial value of the saturation parameter $\phi =
0.1$.

>>>>

We first describe analytic estimates of physical quantities, such as
the stagnation radius $\rs$ and the mass accretion rate, the detailed
derivation of which are given in Appendix \ref{app:rs}.  
%Our results
%are expressed in terms of the SMBH mass $M_{\bullet}$, total heating
%rate $\tilde{v}_{w}$, and wind mass loss parameter $\eta$.

@@

The accretion rate towards the SMBH is given by the total mass loss rate interior to the stagnation radius (eq.~[\ref{eq:q}]), 
\begin{eqnarray}
  \dot{M} &=& 4\pi \int_{0}^{r_{s}}q r^{2}dr = \frac{\eta M_{\star}|_{r_{\rm s}}}{\th} \nonumber \\
&\approx&
  \begin{cases}
    4.5 \times 10^{-5} M_{\bullet,8}^{1.76}
    v_{500}^{-3.8}  \eta_{0.02} \Msun \, \pyear& \text{core} \\
    3.2 \times 10^{-5} M_{\bullet,8}^{1.48} 
    v_{500}^{-2.4}  \eta_{0.02} \Msun \, \pyear  & \text{cusp}, 
  \end{cases}
  \label{eq:mdot_analytic}
\end{eqnarray}
where we have assumed $v_{w} \gg \sigma_0$ by adopting equation
(\ref{eq:stag_simple}) for $r_s$.  The resulting Eddington ratio is
given by
\begin{eqnarray}
\frac{\dot{M}}{\dot{M}_{\rm edd}} &\approx&
  \begin{cases}
    2.0 \times 10^{-5} M_{\bullet,8}^{0.76}
    v_{500}^{-3.8}  \eta_{0.02}   & \text{core}, \\
    1.4 \times 10^{-5} \Mbheight^{0.48} 
    v_{500}^{-2.4}  \eta_{0.02}   & \text{cusp}, 
  \end{cases}
  \label{eq:eddr_analytic}
\end{eqnarray}
where $\dot{M}_{\rm edd} = 2.6M_{\bullet,8}M_{\odot}$ yr$^{-1}$ is the
Eddington accretion rate, assuming a radiative efficiency of ten per
cent.  Note the sensitive dependence of the accretion rate on the wind
heating rate.  As will be discussed later, equation~\eqref{eq:eddr_analytic} is the radial mass inflow rate on relatively large scales and does not include outflows from the SMBH accretion disk (e.g.~\citealt{Blandford&Begelman99}; \citealt{Li+13}), which may significantly reduce the fraction of $\dot{M}$ that actually reaches the SMBH.

@@

Our analytic estimates accurately reproduce the numerical results in
the high heating limit $\zeta \gg 1$ ($v_w \gg \sigma_0$; $r_{\rm s}
\lesssim r_{\rm inf}$).  However, for low heating the stagnation
radius diverges above the analytic estimate, approaching the stellar
break radius $r_b \gg r_{\rm inf}$.  This divergence occurs
approximately when $\zeta < \zeta_c \propto (r_b/r_{\rm
  soi})^{0.5(\Gamma-1)}$ (eq.~[\ref{eq:zetacrit}]), which coincides
with the Bernoulli parameter of the gas decreasing below unity near
the stagnation radius (Appendix \ref{app:be}).  Thus, for small values
of the heating rate (small $\zeta$) the location of the stagnation
radius will vary strongly with the break radius. This explains the
behavior of the three vertically aligned green squares. These are three
cusp ($\Gamma=0.8$) galaxies with t$\Mbh=10^8 \Msun$ and $v_w=300$ km s$^{-1}$
but with different break radii ($\rb=$ 25, 50, and 100 pc from top to
bottom). This divergence of the stagnation radius to large radii
occurs at a higher value of $\zeta$ in core galaxies ($\Gamma = 0.1$),
explaining the behavior of the two core galaxies shown in
Fig.~\ref{fig:stag} as vertically aligned orange triangles.  
%These
%solutions correspond to the same black hole mass ($\Mbh = 10^8 \Msun$)
%and heating rate ($v_w = 600$ km s$^{-1}$) but assume different break
%radii of $r_{b} =$ 25, 50 pc (from top to bottom).

@@

Also note that for $v_{w}=600$ and $1200$ km s$^{-1}$ we have $\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}| \sim t_{\rm cool}/t_{\rm ff}$
near the stagnation radius, but these ratios diverge from each other at low heating ($v_{w} \lesssim 300$ km s$^{-1}$).  This results because equations (\ref{eq:mdot_gas}), (\ref{eq:rhors}) underestimate the true gas density in the case of subsonic flow (weak heating).

Although most of our solutions neglect thermal conduction and
radiative cooling, these effects are explore explicitly in subset of simulations, as marked by dagger symbols in Table \ref{table:models}.  For cases which are far from being thermally
unstable when cooling is neglected (e.g., $M_{\bullet} =
10^{8}M_{\odot}$, $v_{w} = 600$ km s$^{-1}$, $r_{b} = 100$ pc, $\eta =
0.2$, $\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}| = 76$), including radiative cooling has little effect on the key properties of the solution, such as the stagnation radius, mass accretion rate, and the cooling ratio $\dot{q}_{\rm
  heat}/|\dot{q}_{\rm rad}|$.  However, for solutions that are marginally thermally stable, including radiative losses acts to significantly decrease $\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}|$.  We calculated solutions with radiative
cooling for all cusp galaxies with $v_w=300$ km s$^{-1}$ and for all
core galaxies with $v_w=600 $ km s$^{-1}$, each for three different
values of $\eta = 0.02, 0.2, 0.6$, spanning the physical range of
expected mass loss for continuous star formation (Fig.~\ref{fig:vwSources}).  Solutions found to be unstable for all values of $\eta$ are marked in boldface as ``TI" in Table
\ref{table:models}, with the unstable values of $\eta$ provided in the final column.

Including thermal conduction, by contrast, results in only minor changes to our solutions for our fiducial value of the saturation parameter $\phi = 0.1$, consistent with our analytic expectation (eq.~[\ref{eq:conduction}]).
